A Large and Variable Leading Tail of Helium in a Hot Saturn Undergoing Runaway Inflation

### 2023-07-13 paper/paper1/ms.tex
Collapse folders to a flat directory with all figures and tables

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/HAT-P-67b_radius_valley.png}
    \caption{Overview of exoplanet demographics.  The pixel bins reflect the observed density of over 5000 planets accessed from the NASA Exoplanet Archive. Detections of atmospheric escape are common among large planets with strong insolation.  \hatpb is among the most inflated planets known.}
    \label{fig:instellation}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/tf2018_fig2_update2023_HAT.png}
    \caption{Mass-radius trends for inflated hot sub-Saturns and hot Jupiters, with layout following Figure 2 of \citet{2018AJ....155..214T} and updated with NASA Exoplanet Archive confirmed planets.  The trend lines show the mass-radius relationship for equilibrium temperatures of 500, 1000, 1250, 1500, and 2000 K, assuming the mean composition and mean heating efficiency from the original figure.  HAT-P-67~b ($\star$ symbol) stands alone in a region defined by the lack of inflated sub-Saturns and explained by short lifetimes from runaway inflation and mass loss.}
    \label{fig:ThornFortUpdated}
\end{figure}

@@

\begin{deluxetable}{lccrchr}
    \tablewidth{0pc}
    \tabletypesize{\scriptsize}
    \tablecaption{
        HPF Observation Log
        \label{tabHPFLog}
    }
    \tablehead{
        \colhead{UTC}   &
        \colhead{Track} &
        \colhead{Desc.} &
        \colhead{BJD} &
        \colhead{$N_\mathrm{exp}$} &
        \nocolhead{$t_\mathrm{exp}$} &
        \colhead{$\phi$}\\
        \colhead{}   &
        \colhead{} &
        \colhead{} &
        \colhead{2457000.00+} &
        \colhead{} &
        \nocolhead{(s)} &
        \colhead{}
    }
    \startdata
    \input{tables/observation_log.tex}
    \enddata
    \tablecomments{All exposure times were 308.85~s except for observations from 2020-07-22 to 2021-03-31, which were 511.20~s.  Column 2 describes the HET ``track'', restricted to either East (E) or West (W).  Column 5 is the normalized phase $\phi \in(-0.5, 0.5)$, with zero at midtransit.}
\end{deluxetable}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.98\linewidth]{figures/TESS_HAT-P-67b_overview.png}
    \caption{Overview of all available TESS Sectors showing 24 full or partial transits with 34 visits with HPF (vertical gray bars), 7 of which coincide with transits.  The rest of the visits sample out-of-transit phases.  The 8 HPF visits between days 2050 and 2690 are not shown.  Note the large time breaks between some consecutive panels, which correspond roughly to the duration of a TESS sector.}
    \label{fig:TESSoverview}
\end{figure*}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/best_fit_orbit.png}
    \caption{Best fit orbit overlaid on the composite TESS lightcurve.  Revised planet properties are consistent with \citet{2017AJ....153..211Z}, with a systematic shift from a revised \emph{Gaia} DR3 distance.}
    \label{fig:transit}
\end{figure}

@@

\input{tables/revise_props_tbl.tex}

\begin{figure}
    \centering
    \includegraphics[width=0.98\linewidth]{figures/MIST_only_HATP.png}
    \caption{MIST evolutionary model tracks for HAT-P-67.  The $T_\mathrm{eff}$ and luminosity of \hatp (gray circle) are consistent with either the late main sequence or recently evolved subgiant.  The labeled numbers indicate age in Gyr.}
    \label{fig:evolTracks}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/HAT-P-67b_rv.pdf}
    \caption{RV orbit fit including both HPF and Keck HIRES data points.  The RV information content in HPF is much less than in Keck HIRES for this F5 spectral type, and so the joint fit constrains the mass to roughly a Saturn mass.}
    \label{fig:RVfit}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/HAT_P_67b_He_spectrum.png}
    \caption{Overlay of all 152 individual HPF exposures of HAT-P-67, spanning 2020-2022. Variability is seen in the \ion{He}{1} 10833 \AA~ triplet near the vertical orange shaded band but not in the adjacent Si line.  These snapshot spectra were barycentric corrected and continuum flattened with a linear fit to the regions in the blue vertical bands.  Sharp telluric absorption lines have been masked in regions near 10835 and 10837.5 \AA.}
    \label{fig:HPFheliumOverview}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/HAT_P_67b_byCampaign_Spectra.png}
    \caption{Four observing campaigns centered on an in-transit epoch with before-and-after visits typically separated by 1 night. The after-transit spectra tend to show negligible absorption.  Line-of-sight velocity substructure can be seen in the before and during transit epochs.}
    \label{fig:HPFperCampaign}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/HAT_P_67b_Helium10830EW_byCampaign.png}
    \caption{Equivalent Width lightcurve of Helium absorption in HAT-P-67b.  The system exhibits a characteristic absorption leading up to transit, followed by a sharp decline after transit passage.  The EWs were computed in the orange shaded band (10832.3$-$10834.2 \AA) in Figure \ref{fig:HPFheliumOverview}.}
    \label{fig:HPFtimeseries}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/phase_2D_diagram_resid.png}
    \caption{Absorption depth phase scan over the entire HAT-P-67b orbit from 41 visits with HPF.  The planet orbital rest frame is shown as the black sine wave. Gaps in data at 10834-10836 \AA~ arise from telluric masking.  Unmasked telluric lines are perceptible at 10825 \AA.}
    \label{fig:HPFscanResid}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/centroid_overview10833.png}
    %\hskip-3em
    %\subfloat{\includegraphics[width=4cm]{figures/centroid_inset10833.png}}
    \caption{Line centroid positions of the  \ion{He}{1} 10833 \AA~ feature.  A slight blueshift relative to the stellar restframe (vertical line) can be seen, with a weak excess blueshift trend after planet egress. The line-of-sight velocity distribution of 2-3 \AA~ means that the gas distribution exhibits both advancing and receding lobes at all phases other than post-egress.}
    \label{fig:centroids}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/XUV_flux_schematic.png}
    \caption{Synthetic XUV radiation scenarios for \hatpb.  The low and high XUV scenarios correspond to $L_X/L_\mathrm{bol}$ of $10^{-6}$ and $10^{-4}$, where $L_X$ is defined as He 1 $^1$S Ionizing photons that trigger the recombination cascade needed for observing the He 2 $^3$S ionization metastable state.  He 2 $^3$S Ionizing photons depopulate the state and suppress the observability of \ion{He}{1} 10833 \AA.  The XUV flux of HAT-P-67 is uncertain.}
    \label{fig:XUV}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/pwinds_May2020_14000_2e13Mdot_medXUV_formal_rv_rev.png}
    \caption{Simulated 1D Parker wind model of Helium absorption in \hatpb.  The spectrum was generated with a mass loss rate of $2\times10^{13}$ g/s, exosphere gas temperature of $14\,000\;$K, and with an XUV luminosity $L_x/L_\mathrm{bol}=10^{-5}$ intermediate between those shown in Figure \ref{fig:XUV}.  The \texttt{p-winds} model yields the \ion{He}{1} 10833 triplet lines (colored thin lines), with a cumulative feature in black resembling the data, shown as the May 2020 in-transit mean.  Additional 3D velocity dispersion of the gas can explain differences between the line shapes of the model and data.  Overall the 1D models may be too simplistic to represent the inherently 3D structure of the outflow.}
    \label{fig:pwinds}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=0.8\linewidth]{figures/KeplerianShear_v0p5_alt.png}
    \caption{Schematic of Keplerian orbital shear.  Locations outside the Roche lobe experience orbital shear from the Keplerian potential.  A wind launched primarily from the dayside will tend to form a leading tail.}
    \label{fig:KeplerianShear}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/reinflation_MIST2.png}
    \caption{Conceivable evolutionary scenarios for \hatpb.  The planet's equilibrium temperature increases as the stellar luminosity gradually increases on the main sequence.  The second panel from the top shows the corresponding estimate for anomalous heating efficiency from \citet{2018AJ....155..214T}. The Ohmic dissipation model makes predictions for runaway inflation, depending on the incident stellar flux.  The heuristic scaling relation for the inflation timescale predicts vanishingly short inflationary lifetimes in this extreme regime but illustrates the countervailing effects of the top two panels.}
    \label{fig:OhmicInflate}
\end{figure}

@@

\begin{figure*}[t]
    \includegraphics[width=\linewidth]{figures/thorn2023_theory.png}
    \caption{Runaway inflation timescale in the exoplanet mass-radius diagram.  The shading of the left panel shows $\tau_\mathrm{infl}$ from Ohmic dissipation; the right panel shows $\tau_\mathrm{infl}\equiv M/\dot{M}$ from photoionization-driven mass loss.  Individual planets with confident mass and radius detections are shown as small black dots. Helium non-detections are shown in red squares, atmospheric escape detections from any origin in green circles \citep{2022arXiv221116243D}.  Under either scenario, \hatpb~ resides in a sparsely populated region with a short inflationary timescale, making it unstable to runaway evaporation with large mass loss expected.  Photoionization better predicts the depopulation of sources in the upper left low-density region of the diagram, defined by the 0.1 g cm$^{-3}$ iso-density dashed white line.}
    \label{fig:tf2018_theory}
\end{figure*}

@@

\begin{deluxetable*}{lLLCRR}
    \tablewidth{0pc}
    \tablecaption{
        Coarse Inflation Timescales for Other Systems
        \label{tabHeSM}
    }
    \tablehead{
        \colhead{ }   &
        \colhead{ } &
        \colhead{ }   &
        \colhead{ } &
        \colhead{(OD)} & 
        \colhead{(XUV)} \\
        \colhead{Planet}   &
        \colhead{Mass} &
        \colhead{Radius} &
        \colhead{$T_\mathrm{eq}$} &
        \colhead{$\tau_\mathrm{infl}$} &
        \colhead{$\tau_\mathrm{infl}$} \\
        \colhead{}   &
        \colhead{$M_\mathrm{Jup}$} &
        \colhead{$R_\mathrm{Jup}$}   &
        \colhead{K} &
        \colhead{Myr} & 
        \colhead{Myr}
    }
    \startdata
    \input{tables/tab_HeSM.tex}
    \enddata
    \tablecomments{XUV Inflation timescale assumes $L_X/L_\mathrm{bol}=6.3\times10^{-4}$ (Equation 5 of Sanz-Forcada et al. 2011), which is an overestimate for HAT-P-67 and old/slowly-rotating host stars.}
\end{deluxetable*}

@@

\begin{figure}
    \includegraphics[width=0.5\linewidth]{figures/Ca8662_phase_2D_diagram_resid.png}
    \caption{Non-detection of variability in the Calcium IR Triplet line at 8862 \AA.  No other HPF lines appear to show any variability.}
    \label{fig:CaPhaseScan}
\end{figure}

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{HAT-P-67b_radius_valley.png}
    \caption{Overview of exoplanet demographics.  The pixel bins reflect the observed density of over 5000 planets accessed from the NASA Exoplanet Archive. Detections of atmospheric escape are common among large planets with strong insolation.  \hatpb is among the most inflated planets known.}
    \label{fig:instellation}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{tf2018_fig2_update2023_HAT.png}
    \caption{Mass-radius trends for inflated hot sub-Saturns and hot Jupiters, with layout following Figure 2 of \citet{2018AJ....155..214T} and updated with NASA Exoplanet Archive confirmed planets.  The trend lines show the mass-radius relationship for equilibrium temperatures of 500, 1000, 1250, 1500, and 2000 K, assuming the mean composition and mean heating efficiency from the original figure.  HAT-P-67~b ($\star$ symbol) stands alone in a region defined by the lack of inflated sub-Saturns and explained by short lifetimes from runaway inflation and mass loss.}
    \label{fig:ThornFortUpdated}
\end{figure}

@@

\begin{deluxetable}{lccrchr}
    \tablewidth{0pc}
    \tabletypesize{\scriptsize}
    \tablecaption{
        HPF Observation Log
        \label{tabHPFLog}
    }
    \tablehead{
        \colhead{UTC}   &
        \colhead{Track} &
        \colhead{Desc.} &
        \colhead{BJD} &
        \colhead{$N_\mathrm{exp}$} &
        \nocolhead{$t_\mathrm{exp}$} &
        \colhead{$\phi$}\\
        \colhead{}   &
        \colhead{} &
        \colhead{} &
        \colhead{2457000.00+} &
        \colhead{} &
        \nocolhead{(s)} &
        \colhead{}
    }
    \startdata
    \input{observation_log.tex}
    \enddata
    \tablecomments{All exposure times were 308.85~s except for observations from 2020-07-22 to 2021-03-31, which were 511.20~s.  Column 2 describes the HET ``track'', restricted to either East (E) or West (W).  Column 5 is the normalized phase $\phi \in(-0.5, 0.5)$, with zero at midtransit.}
\end{deluxetable}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.98\linewidth]{TESS_HAT-P-67b_overview.png}
    \caption{Overview of all available TESS Sectors showing 24 full or partial transits with 34 visits with HPF (vertical gray bars), 7 of which coincide with transits.  The rest of the visits sample out-of-transit phases.  The 8 HPF visits between days 2050 and 2690 are not shown.  Note the large time breaks between some consecutive panels, which correspond roughly to the duration of a TESS sector.}
    \label{fig:TESSoverview}
\end{figure*}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{best_fit_orbit.png}
    \caption{Best fit orbit overlaid on the composite TESS lightcurve.  Revised planet properties are consistent with \citet{2017AJ....153..211Z}, with a systematic shift from a revised \emph{Gaia} DR3 distance.}
    \label{fig:transit}
\end{figure}

@@

\input{revise_props_tbl.tex}

\begin{figure}
    \centering
    \includegraphics[width=0.98\linewidth]{MIST_only_HATP.png}
    \caption{MIST evolutionary model tracks for HAT-P-67.  The $T_\mathrm{eff}$ and luminosity of \hatp (gray circle) are consistent with either the late main sequence or recently evolved subgiant.  The labeled numbers indicate age in Gyr.}
    \label{fig:evolTracks}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{HAT-P-67b_rv.pdf}
    \caption{RV orbit fit including both HPF and Keck HIRES data points.  The RV information content in HPF is much less than in Keck HIRES for this F5 spectral type, and so the joint fit constrains the mass to roughly a Saturn mass.}
    \label{fig:RVfit}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{HAT_P_67b_He_spectrum.png}
    \caption{Overlay of all 152 individual HPF exposures of HAT-P-67, spanning 2020-2022. Variability is seen in the \ion{He}{1} 10833 \AA~ triplet near the vertical orange shaded band but not in the adjacent Si line.  These snapshot spectra were barycentric corrected and continuum flattened with a linear fit to the regions in the blue vertical bands.  Sharp telluric absorption lines have been masked in regions near 10835 and 10837.5 \AA.}
    \label{fig:HPFheliumOverview}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{HAT_P_67b_byCampaign_Spectra.png}
    \caption{Four observing campaigns centered on an in-transit epoch with before-and-after visits typically separated by 1 night. The after-transit spectra tend to show negligible absorption.  Line-of-sight velocity substructure can be seen in the before and during transit epochs.}
    \label{fig:HPFperCampaign}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{HAT_P_67b_Helium10830EW_byCampaign.png}
    \caption{Equivalent Width lightcurve of Helium absorption in HAT-P-67b.  The system exhibits a characteristic absorption leading up to transit, followed by a sharp decline after transit passage.  The EWs were computed in the orange shaded band (10832.3$-$10834.2 \AA) in Figure \ref{fig:HPFheliumOverview}.}
    \label{fig:HPFtimeseries}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{phase_2D_diagram_resid.png}
    \caption{Absorption depth phase scan over the entire HAT-P-67b orbit from 41 visits with HPF.  The planet orbital rest frame is shown as the black sine wave. Gaps in data at 10834-10836 \AA~ arise from telluric masking.  Unmasked telluric lines are perceptible at 10825 \AA.}
    \label{fig:HPFscanResid}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{centroid_overview10833.png}
    %\hskip-3em
    %\subfloat{\includegraphics[width=4cm]{centroid_inset10833.png}}
    \caption{Line centroid positions of the  \ion{He}{1} 10833 \AA~ feature.  A slight blueshift relative to the stellar restframe (vertical line) can be seen, with a weak excess blueshift trend after planet egress. The line-of-sight velocity distribution of 2-3 \AA~ means that the gas distribution exhibits both advancing and receding lobes at all phases other than post-egress.}
    \label{fig:centroids}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{XUV_flux_schematic.png}
    \caption{Synthetic XUV radiation scenarios for \hatpb.  The low and high XUV scenarios correspond to $L_X/L_\mathrm{bol}$ of $10^{-6}$ and $10^{-4}$, where $L_X$ is defined as He 1 $^1$S Ionizing photons that trigger the recombination cascade needed for observing the He 2 $^3$S ionization metastable state.  He 2 $^3$S Ionizing photons depopulate the state and suppress the observability of \ion{He}{1} 10833 \AA.  The XUV flux of HAT-P-67 is uncertain.}
    \label{fig:XUV}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{pwinds_May2020_14000_2e13Mdot_medXUV_formal_rv_rev.png}
    \caption{Simulated 1D Parker wind model of Helium absorption in \hatpb.  The spectrum was generated with a mass loss rate of $2\times10^{13}$ g/s, exosphere gas temperature of $14\,000\;$K, and with an XUV luminosity $L_x/L_\mathrm{bol}=10^{-5}$ intermediate between those shown in Figure \ref{fig:XUV}.  The \texttt{p-winds} model yields the \ion{He}{1} 10833 triplet lines (colored thin lines), with a cumulative feature in black resembling the data, shown as the May 2020 in-transit mean.  Additional 3D velocity dispersion of the gas can explain differences between the line shapes of the model and data.  Overall the 1D models may be too simplistic to represent the inherently 3D structure of the outflow.}
    \label{fig:pwinds}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=0.8\linewidth]{KeplerianShear_v0p5_alt.png}
    \caption{Schematic of Keplerian orbital shear.  Locations outside the Roche lobe experience orbital shear from the Keplerian potential.  A wind launched primarily from the dayside will tend to form a leading tail.}
    \label{fig:KeplerianShear}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{reinflation_MIST2.png}
    \caption{Conceivable evolutionary scenarios for \hatpb.  The planet's equilibrium temperature increases as the stellar luminosity gradually increases on the main sequence.  The second panel from the top shows the corresponding estimate for anomalous heating efficiency from \citet{2018AJ....155..214T}. The Ohmic dissipation model makes predictions for runaway inflation, depending on the incident stellar flux.  The heuristic scaling relation for the inflation timescale predicts vanishingly short inflationary lifetimes in this extreme regime but illustrates the countervailing effects of the top two panels.}
    \label{fig:OhmicInflate}
\end{figure}

@@

\begin{figure*}[t]
    \includegraphics[width=\linewidth]{thorn2023_theory.png}
    \caption{Runaway inflation timescale in the exoplanet mass-radius diagram.  The shading of the left panel shows $\tau_\mathrm{infl}$ from Ohmic dissipation; the right panel shows $\tau_\mathrm{infl}\equiv M/\dot{M}$ from photoionization-driven mass loss.  Individual planets with confident mass and radius detections are shown as small black dots. Helium non-detections are shown in red squares, atmospheric escape detections from any origin in green circles \citep{2022arXiv221116243D}.  Under either scenario, \hatpb~ resides in a sparsely populated region with a short inflationary timescale, making it unstable to runaway evaporation with large mass loss expected.  Photoionization better predicts the depopulation of sources in the upper left low-density region of the diagram, defined by the 0.1 g cm$^{-3}$ iso-density dashed white line.}
    \label{fig:tf2018_theory}
\end{figure*}

@@

\begin{deluxetable*}{lLLCRR}
    \tablewidth{0pc}
    \tablecaption{
        Coarse Inflation Timescales for Other Systems
        \label{tabHeSM}
    }
    \tablehead{
        \colhead{ }   &
        \colhead{ } &
        \colhead{ }   &
        \colhead{ } &
        \colhead{(OD)} & 
        \colhead{(XUV)} \\
        \colhead{Planet}   &
        \colhead{Mass} &
        \colhead{Radius} &
        \colhead{$T_\mathrm{eq}$} &
        \colhead{$\tau_\mathrm{infl}$} &
        \colhead{$\tau_\mathrm{infl}$} \\
        \colhead{}   &
        \colhead{$M_\mathrm{Jup}$} &
        \colhead{$R_\mathrm{Jup}$}   &
        \colhead{K} &
        \colhead{Myr} & 
        \colhead{Myr}
    }
    \startdata
    \input{tab_HeSM.tex}
    \enddata
    \tablecomments{XUV Inflation timescale assumes $L_X/L_\mathrm{bol}=6.3\times10^{-4}$ (Equation 5 of Sanz-Forcada et al. 2011), which is an overestimate for HAT-P-67 and old/slowly-rotating host stars.}
\end{deluxetable*}

@@

\begin{figure}
    \includegraphics[width=0.5\linewidth]{Ca8662_phase_2D_diagram_resid.png}
    \caption{Non-detection of variability in the Calcium IR Triplet line at 8862 \AA.  No other HPF lines appear to show any variability.}
    \label{fig:CaPhaseScan}
\end{figure}
